---
title: "HAARVIVAC 10X Pre-processing MULTIseqDemux"
author: "Jolie Phan"
output: 
  pdf_document:
    toc: yes
    toc_depth: 4
  html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
date: "version `r format(Sys.time(), '%B %d, %Y')`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r}
library(dplyr)
library(Seurat)
library(patchwork)
library(hdf5r)
library(stringr)
library(ggplot2)
library(cowplot)
```

# Load RNA, ADT, and HTO data from .h5 files 

```{r}
save_output <- FALSE
addTitleToMultiPlot <- function(plotGrid, title, relHeights = c(0.1, 1)) {
plot_grid(ggdraw() + draw_label(title), plotGrid, ncol = 1, rel_heights = relHeights)
}

s1_dat <- Read10X_h5(here::here("data/S1/filtered_feature_bc_matrix.h5"))
s2_dat <- Read10X_h5(here::here("data/S2/filtered_feature_bc_matrix.h5"))

if(!dir.exists(here::here("processed_data"))) {
  cat(sprintf("Creating folder %s\n", here::here("processed_data")))
  dir.create(here::here("processed_data"), recursive = T)
}

# Shorten ADT and HTO names
rownames(s1_dat$`Antibody Capture`) <- rownames(s1_dat$`Antibody Capture`) %>%
  str_replace("_TotalSeqCHashing", "") %>%
  str_replace("_TotalSeqC", "")

rownames(s2_dat$`Antibody Capture`) <- rownames(s2_dat$`Antibody Capture`) %>%
  str_replace("_TotalSeqCHashing", "") %>%
  str_replace("_TotalSeqC", "")

# Extract matrices of ADT and HTO info 
s1_ADT <- !grepl("Hashtag", s1_dat$`Antibody Capture`@Dimnames[[1]])
s1_ADT_counts <- s1_dat$`Antibody Capture`[s1_ADT,]
s1_HTO <- grepl("Hashtag", s1_dat$`Antibody Capture`@Dimnames[[1]])
s1_HTO_counts <- s1_dat$`Antibody Capture`[s1_HTO,]

s2_ADT <- !grepl("Hashtag", s2_dat$`Antibody Capture`@Dimnames[[1]])
s2_ADT_counts <- s2_dat$`Antibody Capture`[s2_ADT,]
s2_HTO <- grepl("Hashtag", s2_dat$`Antibody Capture`@Dimnames[[1]])
s2_HTO_counts <- s2_dat$`Antibody Capture`[s2_HTO,]
```

# Set up Seurat object 

```{r}
s1 <- CreateSeuratObject(counts = s1_dat$`Gene Expression`, project = "HAARVIVAC_10X_S1", min.cells = 3)
s1[["ADT"]] <-  CreateAssayObject(counts = s1_ADT_counts)
s1[["HTO"]] <-  CreateAssayObject(counts = s1_HTO_counts)
Assays(s1)

s2 <- CreateSeuratObject(counts = s2_dat$`Gene Expression`, project = "HAARVIVAC_10X_S2", min.cells = 3)
s2[["ADT"]] <-  CreateAssayObject(counts = s2_ADT_counts)
s2[["HTO"]] <-  CreateAssayObject(counts = s2_HTO_counts)
Assays(s2)
```

# Demultiplex cells based on HTO enrichment
MULTIseqDemux() demultiplexes samples based on the classification method from MULTI-seq. 

The "autoThresh" argument decides whether to perform automated threshold finding to define the best quantile. 

```{r, fig.width=12, fig.height=10}
# Normalize HTO data using centered log-ratio (CLR) transformation 
s1 <- NormalizeData(s1, assay = "HTO", normalization.method = "CLR")
s2 <- NormalizeData(s2, assay = "HTO", normalization.method = "CLR")

# Demultiplex
s1_dm <- MULTIseqDemux(s1, assay = "HTO", autoThresh = TRUE)
s2_dm <- MULTIseqDemux(s2, assay = "HTO", autoThresh = TRUE)

s1_start_count <- length(s1_dm$orig.ident)
s2_start_count <- length(s2_dm$orig.ident)

# Assign global classification
i1 <- grepl("Hashtag", s1_dm@meta.data$MULTI_ID)
i2 <- grepl("Doublet", s1_dm@meta.data$MULTI_ID)
i3 <- grepl("Negative", s1_dm@meta.data$MULTI_ID)
s1_dm@meta.data$MULTI_ID_global <- NULL
s1_dm@meta.data$MULTI_ID_global[i1] <- "Singlet"
s1_dm@meta.data$MULTI_ID_global[i2] <- "Doublet"
s1_dm@meta.data$MULTI_ID_global[i3] <- "Negative"
s1_dm@meta.data$MULTI_ID_global <- factor(s1_dm@meta.data$MULTI_ID_global, levels = c("Doublet", "Negative", "Singlet"))

i1 <- grepl("Hashtag", s2_dm@meta.data$MULTI_ID)
i2 <- grepl("Doublet", s2_dm@meta.data$MULTI_ID)
i3 <- grepl("Negative", s2_dm@meta.data$MULTI_ID)
s2_dm@meta.data$MULTI_ID_global <- NULL
s2_dm@meta.data$MULTI_ID_global[i1] <- "Singlet"
s2_dm@meta.data$MULTI_ID_global[i2] <- "Doublet"
s2_dm@meta.data$MULTI_ID_global[i3] <- "Negative"
s2_dm@meta.data$MULTI_ID_global <- factor(s2_dm@meta.data$MULTI_ID_global, levels = c("Doublet", "Negative", "Singlet"))

# Global classification results
table(s1_dm$MULTI_ID_global)
table(s2_dm$MULTI_ID_global)

# Assign maxID (name of hashtag with the highest signal)
s1_maxID <- as.data.frame(GetAssayData(s1_dm, slot = "data", assay = "HTO")) %>%
  tibble::rownames_to_column(var = "Hashtag") %>%
  tidyr::pivot_longer(-Hashtag, names_to = "Name", values_to = "Data") %>%
  tidyr::pivot_wider(names_from = "Hashtag", values_from = "Data") %>%
  tibble::column_to_rownames(var = "Name") %>%
  mutate(MULTI_maxID = purrr::pmap(across(everything()), ~ names(c(...)[which.max(c(...))]))) %>%
  select(MULTI_maxID)

s1_dm@meta.data <- s1_dm@meta.data %>%
  merge(s1_maxID, by = 0) %>%
  tibble::column_to_rownames(var = "Row.names")

s2_maxID <- as.data.frame(GetAssayData(s2_dm, slot = "data", assay = "HTO")) %>%
  tibble::rownames_to_column(var = "Hashtag") %>%
  tidyr::pivot_longer(-Hashtag, names_to = "Name", values_to = "Data") %>%
  tidyr::pivot_wider(names_from = "Hashtag", values_from = "Data") %>%
  tibble::column_to_rownames(var = "Name") %>%
  mutate(MULTI_maxID = purrr::pmap(across(everything()), ~ names(c(...)[which.max(c(...))]))) %>%
  select(MULTI_maxID)

s2_dm@meta.data <- s2_dm@meta.data %>%
  merge(s2_maxID, by = 0) %>%
  tibble::column_to_rownames(var = "Row.names")

# Assign PTIDs and infection status
i1 <- grepl("Hashtag-1", s1_dm@meta.data$MULTI_ID)
i2 <- grepl("Hashtag-2", s1_dm@meta.data$MULTI_ID)
i3 <- grepl("Hashtag-3", s1_dm@meta.data$MULTI_ID) 
i4 <- grepl("Hashtag-4", s1_dm@meta.data$MULTI_ID)
i5 <- grepl("Hashtag-5", s1_dm@meta.data$MULTI_ID)
i6 <- grepl("Hashtag-6", s1_dm@meta.data$MULTI_ID)
i7 <- grepl("Hashtag-7", s1_dm@meta.data$MULTI_ID) 
i8 <- grepl("Hashtag-8", s1_dm@meta.data$MULTI_ID) 
i9 <- grepl("Doublet", s1_dm@meta.data$MULTI_ID) 
i10 <- grepl("Negative", s1_dm@meta.data$MULTI_ID) 
s1_dm@meta.data$PTID <- NULL
s1_dm@meta.data$PTID[i1] <- "50H"
s1_dm@meta.data$PTID[i2] <- "53H"
s1_dm@meta.data$PTID[i3] <- "54H"
s1_dm@meta.data$PTID[i4] <- "55H"
s1_dm@meta.data$PTID[i5] <- "120C"
s1_dm@meta.data$PTID[i6] <- "149C"
s1_dm@meta.data$PTID[i7] <- "194C"
s1_dm@meta.data$PTID[i8] <- "239C"
s1_dm@meta.data$PTID[i9] <- "Doublet"
s1_dm@meta.data$PTID[i10] <- "Negative"

i11 <- grepl("H", s1_dm@meta.data$PTID) 
i12 <- grepl("C", s1_dm@meta.data$PTID)
s1_dm@meta.data$Group <- NULL
s1_dm@meta.data$Group[i11] <- "Naive"
s1_dm@meta.data$Group[i12] <- "Conv"
s1_dm@meta.data$Group[i9] <- "Doublet"
s1_dm@meta.data$Group[i10] <- "Negative"
s1_dm@meta.data$Group <- factor(s1_dm@meta.data$Group, levels = c("Conv", "Naive", "Doublet", "Negative"))

i1 <- grepl("Hashtag-1", s2_dm@meta.data$MULTI_ID)
i2 <- grepl("Hashtag-2", s2_dm@meta.data$MULTI_ID)
i3 <- grepl("Hashtag-3", s2_dm@meta.data$MULTI_ID) 
i4 <- grepl("Hashtag-4", s2_dm@meta.data$MULTI_ID)
i5 <- grepl("Hashtag-5", s2_dm@meta.data$MULTI_ID)
i6 <- grepl("Hashtag-6", s2_dm@meta.data$MULTI_ID)
i7 <- grepl("Hashtag-7", s2_dm@meta.data$MULTI_ID) 
i8 <- grepl("Hashtag-8", s2_dm@meta.data$MULTI_ID) 
i9 <- grepl("Doublet", s2_dm@meta.data$MULTI_ID) 
i10 <- grepl("Negative", s2_dm@meta.data$MULTI_ID) 
s2_dm@meta.data$PTID <- NULL
s2_dm@meta.data$PTID[i1] <- "50H"
s2_dm@meta.data$PTID[i2] <- "53H"
s2_dm@meta.data$PTID[i3] <- "54H"
s2_dm@meta.data$PTID[i4] <- "55H"
s2_dm@meta.data$PTID[i5] <- "120C"
s2_dm@meta.data$PTID[i6] <- "149C"
s2_dm@meta.data$PTID[i7] <- "194C"
s2_dm@meta.data$PTID[i8] <- "239C"
s2_dm@meta.data$PTID[i9] <- "Doublet"
s2_dm@meta.data$PTID[i10] <- "Negative"

i11 <- grepl("H", s2_dm@meta.data$PTID) 
i12 <- grepl("C", s2_dm@meta.data$PTID)
s2_dm@meta.data$Group <- NULL
s2_dm@meta.data$Group[i11] <- "Naive"
s2_dm@meta.data$Group[i12] <- "Conv"
s2_dm@meta.data$Group[i9] <- "Doublet"
s2_dm@meta.data$Group[i10] <- "Negative"
s2_dm@meta.data$Group <- factor(s2_dm@meta.data$Group, levels = c("Conv", "Naive", "Doublet", "Negative"))

# if(save_output) {
#   saveRDS(s1_dm, file = here::here("out/20211118_s1_mdm_all.rds"))
#   saveRDS(s2_dm, file = here::here("out/20211118_s2_mdm_all.rds"))  
# }
```

## HTO enrichment 

```{r, fig.width=12, fig.height=10}
Idents(s1_dm) <- "MULTI_maxID"
RidgePlot(s1_dm, assay = "HTO", features = rownames(s1_dm[["HTO"]]), ncol = 3)

Idents(s2_dm) <- "MULTI_maxID"
RidgePlot(s2_dm, assay = "HTO", features = rownames(s2_dm[["HTO"]]), ncol = 3)
```

## Number of UMIs for singlets, doublets, and negative cells

```{r, fig.width=12, fig.height=10}
Idents(s1_dm) <- "MULTI_ID_global"
VlnPlot(s1_dm, features = "nCount_RNA", pt.size = 0.1, log = TRUE)

Idents(s2_dm) <- "MULTI_ID_global"
VlnPlot(s2_dm, features = "nCount_RNA", pt.size = 0.1, log = TRUE)
```

```{r}
# Remove negative cells
s1_dm_subset <- subset(s1_dm, idents = "Negative", invert = TRUE)
s2_dm_subset <- subset(s2_dm, idents = "Negative", invert = TRUE)
```

## tSNE embedding of the HTO data

```{r, fig.width=12, fig.height=10}
DefaultAssay(s1_dm_subset) <- "HTO"
s1_dm_subset <- ScaleData(s1_dm_subset, features = rownames(s1_dm_subset),
                                 verbose = FALSE)
s1_dm_subset <- RunPCA(s1_dm_subset, features = rownames(s1_dm_subset), approx = FALSE)
s1_dm_subset <- RunTSNE(s1_dm_subset, dims = 1:8, perplexity = 100)
DimPlot(s1_dm_subset) + ggtitle("Spike 1")

DefaultAssay(s2_dm_subset) <- "HTO"
s2_dm_subset <- ScaleData(s2_dm_subset, features = rownames(s2_dm_subset),
                          verbose = FALSE)
s2_dm_subset <- RunPCA(s2_dm_subset, features = rownames(s2_dm_subset), approx = FALSE)
s2_dm_subset <- RunTSNE(s2_dm_subset, dims = 1:8, perplexity = 100)
DimPlot(s2_dm_subset) + ggtitle("Spike 2")
```

Are the # of cells per Group and PTID balanced?

```{r, fig.width=12, fig.height=10}
Idents(s1_dm_subset) <- "PTID"
DimPlot(s1_dm_subset) + ggtitle("Spike 1")

Idents(s2_dm_subset) <- "PTID"
DimPlot(s2_dm_subset) + ggtitle("Spike 2")

Idents(s1_dm_subset) <- "Group"
DimPlot(s1_dm_subset) + ggtitle("Spike 1")

Idents(s2_dm_subset) <- "Group"
DimPlot(s2_dm_subset) + ggtitle("Spike 2")
```

```{r}
# Remove doublets and negative cells
s1_dm_singlets <- subset(s1_dm, idents = "Singlet")
s2_dm_singlets <- subset(s2_dm, idents = "Singlet")

s1_singlet_count <- length(s1_dm_singlets$orig.ident)
s2_singlet_count <- length(s2_dm_singlets$orig.ident)
```

# Analyze ADT data

```{r}
DefaultAssay(s1_dm_singlets) <- "ADT"
rownames(s1_dm_singlets)
```

## ADT enrichment
There are 35 markers and 3 IgG negative controls total. 

```{r, fig.width=12, fig.height=10}
s1_dm_singlets <- NormalizeData(s1_dm_singlets, normalization.method = "CLR", margin = 2, assay = "ADT")
RidgePlot(s1_dm_singlets, assay = "ADT", features = c("CD3", "CD4", "CD8a", "GD", "CD62L", "TRAV1-2", "CD28", "CD127", "CD95", "Mouse-IgG1-Isotype-Control"), ncol = 2) %>%
  addTitleToMultiPlot("Spike 1")

s2_dm_singlets <- NormalizeData(s2_dm_singlets, normalization.method = "CLR", margin = 2, assay = "ADT")
RidgePlot(s2_dm_singlets, assay = "ADT", features = c("CD3", "CD4", "CD8a", "GD", "CD62L", "TRAV1-2", "CD28", "CD127", "CD95", "Mouse-IgG1-Isotype-Control"), ncol = 2) %>%
  addTitleToMultiPlot("Spike 2")
```

# Analyze RNA data

```{r}
DefaultAssay(s1_dm_singlets) <- "RNA"
DefaultAssay(s2_dm_singlets) <- "RNA"
```

## QC
Filter cells that have unique feature counts >200 and mitochondrial counts <5%

```{r, fig.width=12, fig.height=10}
s1_dm_singlets[["percent.mt"]] <- PercentageFeatureSet(s1_dm_singlets, pattern = "^MT-")
VlnPlot(s1_dm_singlets, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3) %>%
  addTitleToMultiPlot("Spike 1")
s1_dm_singlets <- subset(s1_dm_singlets, subset = nFeature_RNA > 200)
s1_200_gene_count <- length(s1_dm_singlets$orig.ident)

s1_dm_singlets <- subset(s1_dm_singlets, subset = percent.mt < 5)
s1_5_mt_count <- length(s1_dm_singlets$orig.ident)

s2_dm_singlets[["percent.mt"]] <- PercentageFeatureSet(s2_dm_singlets, pattern = "^MT-")
VlnPlot(s2_dm_singlets, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3) %>%
  addTitleToMultiPlot("Spike 2")
s2_dm_singlets <- subset(s2_dm_singlets, subset = nFeature_RNA > 200)
s2_200_gene_count <- length(s2_dm_singlets$orig.ident)

s2_dm_singlets <- subset(s2_dm_singlets, subset = percent.mt < 5)
s2_5_mt_count <- length(s2_dm_singlets$orig.ident)
```

Make table of QC cell counts.

```{r}
s1_qc_counts <- data.frame(s1_start_count, s1_singlet_count, s1_200_gene_count, s1_5_mt_count)

s2_qc_counts <- data.frame(s2_start_count, s2_singlet_count, s2_200_gene_count, s2_5_mt_count)

saveRDS(s1_qc_counts, file = here::here("out/s1_qc_counts.rds"))
saveRDS(s2_qc_counts, file = here::here("out/s2_qc_counts.rds"))
```

## Perform visualization and clustering steps

```{r, fig.width=12, fig.height=10}
s1_dm_singlets <- NormalizeData(s1_dm_singlets)
s1_dm_singlets <- FindVariableFeatures(s1_dm_singlets)
s1_dm_singlets <- ScaleData(s1_dm_singlets)
s1_dm_singlets <- RunPCA(s1_dm_singlets, verbose = FALSE)
ElbowPlot(s1_dm_singlets)
s1_dm_singlets <- FindNeighbors(s1_dm_singlets, dims = 1:10)
s1_dm_singlets <- FindClusters(s1_dm_singlets, resolution = 0.8, verbose = FALSE)
s1_dm_singlets <- RunUMAP(s1_dm_singlets, dims = 1:10)
DimPlot(s1_dm_singlets, label = TRUE) + ggtitle("Spike 1")

s2_dm_singlets <- NormalizeData(s2_dm_singlets)
s2_dm_singlets <- FindVariableFeatures(s2_dm_singlets)
s2_dm_singlets <- ScaleData(s2_dm_singlets)
s2_dm_singlets <- RunPCA(s2_dm_singlets, verbose = FALSE)
ElbowPlot(s2_dm_singlets)
s2_dm_singlets <- FindNeighbors(s2_dm_singlets, dims = 1:10)
s2_dm_singlets <- FindClusters(s2_dm_singlets, resolution = 0.8, verbose = FALSE)
s2_dm_singlets <- RunUMAP(s2_dm_singlets, dims = 1:10)
DimPlot(s2_dm_singlets, label = TRUE) + ggtitle("Spike 2")

if(save_output) {
  saveRDS(s1_dm_singlets, file = here::here("processed_data/20211118_s1_mdm_singlets.rds"))
  saveRDS(s2_dm_singlets, file = here::here("processed_data/20211118_s2_mdm_singlets.rds"))  
}
```
