---
title: "HAARVIVAC 10X Weighted Nearest Neighbor Analysis"
author: "Jolie Phan"
output: 
  html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
    toc_depth: 4
date: "version `r format(Sys.time(), '%B %d, %Y')`"
editor_options: 
  chunk_output_type: console
---

Perform Weighted Nearest Neighbor Analysis on T cells.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r}
library(dplyr)
library(stringr)
library(Seurat)
library(cowplot)
library(ggplot2)
library(ggpubr)
library(ggbeeswarm)
```

# Load data

```{r}
save_output <- TRUE
recluster <- TRUE
if(recluster) {
  s1 <- readRDS(file = here::here("processed_data/20220323_s1_tcells_only.rds"))
  s2 <- readRDS(file = here::here("processed_data/20220323_s2_tcells_only.rds"))
} else {
  s1 <- readRDS(file = here::here("processed_data/20220323_s1_tcells_with_wnn.rds"))
  s2 <- readRDS(file = here::here("processed_data/20220323_s2_tcells_with_wnn.rds"))
}

# Set colors
s1_adt_colors <- c("CD4 T cells" = "#66C2A5",
                   "CD8 T cells" = "#FC8D62")

s2_adt_colors <- c("CD4 T cells" = "#1B9E77",
                   "CD8 T cells" = "#D95F02")
```

# Preprocessing and dimensional reduction

After subsetting, FindVariableFeatures, ScaleData, and RunPCA should be performed again.

## Spike 1

```{r}
if(recluster) {
 # RNA
  DefaultAssay(s1) <- "RNA"
  s1 <- FindVariableFeatures(s1) %>%
    ScaleData() %>%
    RunPCA(verbose = FALSE, reduction.key = "PCtcell_", reduction.name = "pca_tcell")

  # ADT
  DefaultAssay(s1) <- "ADT_bc"
  VariableFeatures(s1) <- rownames(s1[["ADT_bc"]])
  s1 <- ScaleData(s1) %>%
    RunPCA(reduction.name = "apca_tcell", reduction.key = "APCtcell_") 
}
```

## Spike 2

```{r}
if(recluster) {
  # RNA
  DefaultAssay(s2) <- "RNA"
  s2 <- FindVariableFeatures(s2) %>%
    ScaleData() %>%
    RunPCA(verbose = FALSE, reduction.key = "PCtcell_", reduction.name = "pca_tcell")

  # ADT
  DefaultAssay(s2) <- "ADT_bc"
  VariableFeatures(s2) <- rownames(s2[["ADT_bc"]])
  s2 <- ScaleData(s2) %>%
    RunPCA(reduction.name = "apca_tcell", reduction.key = "APCtcell_") 
}
```

# Identify multimodal neighbors

## Spike 1

```{r, fig.width=6, fig.height=5}
DefaultAssay(s1) <- "RNA"
ElbowPlot(s1, reduction = "pca_tcell")
```

```{r, fig.width=6, fig.height=5}
DefaultAssay(s1) <- "ADT_bc"
ElbowPlot(s1, reduction = "apca_tcell")
```

```{r}
if(recluster) {
  s1 <- FindMultiModalNeighbors(s1, reduction.list = list("pca_tcell", "apca_tcell"),
                                dims.list = list(1:8, 1:10),
                                modality.weight.name = list("RNA.weight", "ADT.weight"))
} 
```

## Spike 2

```{r, fig.width=6, fig.height=5}
DefaultAssay(s2) <- "RNA"
ElbowPlot(s2, reduction = "pca_tcell")
```

```{r, fig.width=6, fig.height=5}
DefaultAssay(s2) <- "ADT_bc"
ElbowPlot(s2, reduction = "apca_tcell")
```

```{r, fig.width=6, fig.height=5}
if(recluster) {
  s2 <- FindMultiModalNeighbors(s2, reduction.list = list("pca_tcell", "apca_tcell"),
                                dims.list = list(1:8, 1:10),
                                modality.weight.name = list("RNA.weight", "ADT.weight"))
} 
```

# Run UMAP on a weighted combo of RNA and ADT data

## Spike 1

```{r, fig.width=10, fig.height=4}
if(recluster) {
  s1 <- RunUMAP(s1, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")
  s1 <- FindClusters(s1, graph.name = "wsnn", algorithm = 3, resolution = 1, verbose = FALSE)
}

p1 <- DimPlot(s1, reduction = 'wnn.umap', label = TRUE, repel = TRUE, label.size = 4, pt.size = 0.7) + NoLegend() + ggtitle("Spike 1")
p2 <- DimPlot(s1, reduction = 'wnn.umap', group.by = 'adt_clusters', label = FALSE, repel = TRUE, label.size = 4, cols = s1_adt_colors, pt.size = 0.7)
p1 + p2
```

## Spike 2

```{r, fig.width=10, fig.height=4}
if(recluster){
  s2 <- RunUMAP(s2, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")
  s2 <- FindClusters(s2, graph.name = "wsnn", algorithm = 3, resolution = 1, verbose = FALSE)
}

p3 <- DimPlot(s2, reduction = 'wnn.umap', label = TRUE, repel = TRUE, label.size = 4, pt.size = 0.7) + NoLegend() + ggtitle("Spike 2")
p4 <- DimPlot(s2, reduction = 'wnn.umap', group.by = 'adt_clusters', label = FALSE, repel = TRUE, label.size = 4, cols = s2_adt_colors, pt.size = 0.7) 
p3 + p4
```

## Spike 1 and 2 side by side 

```{r, fig.width=9, fig.height=4}
p1 + p3
```

# Visualize ADT expression

Visualize the expression of a few lineage-defining ADT markers.

## Spike 1

```{r, fig.width=8, fig.height=6}
DefaultAssay(s1) <- "ADT_bc"
p5 <- FeaturePlot(s1, features = c("adt_CD3", "adt_CD4", "adt_CD8a", "adt_CD62L", "adt_CD25", "adt_HLA-DR"), reduction = "wnn.umap", ncol = 3)
p5
```

## Spike 2

```{r, fig.width=8, fig.height=6}
DefaultAssay(s2) <- "ADT_bc"
p6 <- FeaturePlot(s2, features = c("adt_CD3", "adt_CD4", "adt_CD8a", "adt_CD62L", "adt_CD25", "adt_HLA-DR"), reduction = "wnn.umap", ncol = 3)
p6
```

# Visualize the modality weights that were learned for each cell

## Spike 1

```{r}
VlnPlot(s1, features = "RNA.weight", sort = TRUE, pt.size = 0.1) + NoLegend()

VlnPlot(s1, features = "ADT.weight", sort = TRUE, pt.size = 0.1) + NoLegend()
```

```{r}
VlnPlot(s1, features = "RNA.weight", group.by = "adt_clusters", sort = TRUE, pt.size = 0.1, cols = s1_adt_colors) + NoLegend()

VlnPlot(s1, features = "ADT.weight", group.by = "adt_clusters", sort = TRUE, pt.size = 0.1, cols = s1_adt_colors) + NoLegend()
```

## Spike 2

```{r}
VlnPlot(s2, features = "RNA.weight", sort = TRUE, pt.size = 0.1) + NoLegend()

VlnPlot(s2, features = "ADT.weight", sort = TRUE, pt.size = 0.1) + NoLegend()
```

```{r}
VlnPlot(s2, features = "RNA.weight", group.by = "adt_clusters", sort = TRUE, pt.size = 0.1, cols = s2_adt_colors) + NoLegend()

VlnPlot(s2, features = "ADT.weight", group.by = "adt_clusters", sort = TRUE, pt.size = 0.1, cols = s2_adt_colors) + NoLegend()
```

# Dotplots

Proportion of cells falling within each cluster per PTID.

```{r}
make_dotplots <- function(data, groups_to_compare, stim_for_title) {
  clust_prop <- table(data$wsnn_res.1, data$Group, data$PTID) %>%
    as.data.frame() %>%
    filter(Freq != 0) %>%
    filter(Var2 %in% groups_to_compare) %>%
    group_by(Var3) %>%
    mutate_at(.vars = "Freq", list(~./sum(.))) %>%
    mutate(Pct = Freq*100) %>%
    select(-Freq) %>%
    dplyr::rename(Cluster = Var1) %>%
    dplyr::rename(Group = Var2) %>%
    dplyr::rename(PTID = Var3)
  
  clust_dotplot <- ggplot(clust_prop, aes(x = Group, y = Pct, fill = Group)) +
    geom_signif(comparisons = list(c("Conv", "Naive")),
              map_signif_level = TRUE,
              test = "wilcox.test",
              test.args = list(paired = FALSE),
              textsize = 3,
              vjust = -0.2) +
    stat_summary(fun = median, geom = "crossbar", width = 0.3) +
    geom_quasirandom(aes(fill = Group), pch = 21,
                   width = 0.1, size = 1.5) +
    facet_wrap(~Cluster, scale="free") +
    ylab("% of T cells") +
    theme(text = element_text(family = "Arial"),
          axis.title.x = element_blank(),
          legend.title = element_blank(),
          plot.title = element_text(hjust = 0.5),
          plot.margin = margin(0.3, 0.2, 0.1, 0.2, "cm"),
          panel.grid = element_blank()) +
    scale_fill_manual(values = c(Conv = "#E31A1C", Naive = "#1F78B4")) +
    theme(legend.position="none") +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
    ggtitle(stim_for_title) 
}
```

## Spike 1

```{r, fig.width=6, fig.height=7}
s1_clust_dotplot <- make_dotplots(s1, groups_to_compare = c("Conv", "Naive"), stim_for_title = "Spike 1")

s1_clust_dotplot
```

## Spike 2

```{r, fig.width=6, fig.height=7}
s2_clust_dotplot <- make_dotplots(s2, groups_to_compare = c("Conv", "Naive"), stim_for_title = "Spike 2")

s2_clust_dotplot
```

# Save output

```{r}
if(save_output) {
  saveRDS(s1, file = here::here("processed_data/20220323_s1_tcells_with_wnn.rds"))
  saveRDS(s2, file = here::here("processed_data/20220323_s2_tcells_with_wnn.rds"))
}
```
