---
title: "HAARVIVAC 10X Mixed Model"
author: "Jolie Phan"
output: 
  html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
    toc_depth: 4
date: "version `r format(Sys.time(), '%B %d, %Y')`"
editor_options: 
  chunk_output_type: console
---

The purpose of this analysis is to test associations between cluster and infection status, correcting for technical factors and other confounding covariates.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r, message=FALSE}
library(lme4)
library(readxl)
library(dplyr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(ggh4x)
```

# Load data 

```{r}
save_output <- TRUE
rerun_masc <- FALSE

s1 <- readRDS(file = here::here("processed_data/20220413_s1_tcells_with_wnn_names.rds")) 
s2 <- readRDS(file = here::here("processed_data/20220413_s2_tcells_with_wnn_names.rds"))
metadata <- read_excel(here::here("data/2022_HAARVIVAC_SuppTable1.xlsx"), sheet = 1)

# Subset data to drop Cluster 14 from S1 and Cluster 11 from S2
s1 <- subset(s1, subset = wsnn_res.1 != "14")
s2 <- subset(s2, subset = wsnn_res.1 != "11")

if(!dir.exists(here::here("out/mixed_models"))) {
    cat(sprintf("Creating folder %s\n", here::here("out/mixed_models")))
    dir.create(here::here("out/mixed_models"), recursive = T)
}
```

# Mixed model

Covariate of interest: Group

Random effects: PTID

Fixed effects: nCount_RNA, nCount_ADT, percent.mt, sex, age

First, get the necessary demographic information.

```{r}
metadata <- metadata %>%
  select(PTID, Age, Sex)
```

## Spike 1

```{r, fig.width=9, fig.height=6}
# Get data frame 
s1_df <- s1@meta.data %>%
  select(PTID, nCount_RNA, nCount_ADT, percent.mt, Group, wsnn_res.1)
s1_df <- inner_join(s1_df, metadata, by = "PTID")
s1_df$Group <- factor(s1_df$Group, levels = c("Naive", "Conv"))
s1_df$Sex <- factor(s1_df$Sex, levels = c("Female", "Male"))

# Rescale UMIs
s1_df$nCount_RNA <- scale(log2(s1_df$nCount_RNA))
s1_df$nCount_ADT <- scale(log2(s1_df$nCount_ADT))

# Create design matrix
s1_cluster <- as.character(s1_df$wsnn_res.1)
s1_designmat <- model.matrix(~ s1_cluster + 0, data.frame(cluster = s1_cluster))
s1_df <- cbind(s1_designmat, s1_df)

# Set variables
contrast <- "Group"
random <- "PTID"
fixed <- c("nCount_RNA", "nCount_ADT", "percent.mt", "Age", "Sex")

# Create base model formula
formula_rhs <- paste0(c(paste0(fixed, collapse = " + "),
                      paste0("(1|", random, ")", collapse = " + ")),
                    collapse = " + ")

# Initialize list to store model objects for each cluster
s1_cluster_models <- vector(mode = "list",
                         length = length(attributes(s1_designmat)$dimnames[[2]]))
names(s1_cluster_models) <- attributes(s1_designmat)$dimnames[[2]]

if(rerun_masc) {
  # Run mixed-effects models for each cluster
  system.time({
    for (i in seq_along(colnames(s1_designmat))) {
      test_cluster <- colnames(s1_designmat)[i]
      null_formula <- as.formula(paste0(c(paste0(test_cluster, " ~ 1 + "),
                                          formula_rhs), collapse = ""))
      full_formula <- as.formula(paste0(c(paste0(test_cluster, " ~ ", contrast, " + "),
                                          formula_rhs), collapse = ""))
      # Run null and full mixed-effects models
      null_model <- lme4::glmer(formula = null_formula, data = s1_df,
                                family = binomial, nAGQ = 1, verbose = 0,
                                control = glmerControl(optimizer = "bobyqa"))
      full_model <- lme4::glmer(formula = full_formula, data = s1_df,
                                family = binomial, nAGQ = 1, verbose = 0,
                                control = glmerControl(optimizer = "bobyqa"))
      
      # Likelihood ratio test to test fixed effects
      model_lrt <- anova(null_model, full_model)
  
      # Calculate confidence intervals for contrast term beta
      contrast_lvl2 <- paste0(contrast, levels(s1_df[[contrast]])[2])
      contrast_ci <- confint.merMod(full_model, method = "Wald",
                                  parm = contrast_lvl2)
  
      # Save model objects to list
      s1_cluster_models[[i]]$null_model <- null_model
      s1_cluster_models[[i]]$full_model <- full_model
      s1_cluster_models[[i]]$model_lrt <- model_lrt
      s1_cluster_models[[i]]$confint <- contrast_ci
      }  
    })
  
  # Organize results into output dataframe
  s1_output <- data.frame(cluster = attributes(s1_designmat)$dimnames[[2]],
                          size = colSums(s1_designmat))
  s1_output$model.pvalue <- sapply(s1_cluster_models, function(x) x$model_lrt[["Pr(>Chisq)"]][2])
  s1_output$cluster <- s1_output$cluster %>%
    str_replace("s1_cluster", "")
  
  # Get odds ratio for Conv vs. Naive
  s1_output[[paste(contrast_lvl2, "OR", sep = ".")]] <- sapply(s1_cluster_models, function(x) exp(fixef(x$full)[[contrast_lvl2]]))
  s1_output[[paste(contrast_lvl2, "OR", "95pct.ci.lower", sep = ".")]] <- sapply(s1_cluster_models, function(x) exp(x$confint[contrast_lvl2, "2.5 %"]))
  s1_output[[paste(contrast_lvl2, "OR", "95pct.ci.upper", sep = ".")]] <- sapply(s1_cluster_models, function(x) exp(x$confint[contrast_lvl2, "97.5 %"]))
  } else {
    s1_output <- readRDS(file = here::here("out/mixed_models/S1_masc_output.rds"))
}

# Select significant points to highlight 
s1_highlight <- s1_output %>%
  dplyr::filter(-log10(model.pvalue) > -log10(0.05))

# Visualize associations between infection status (Group) and WNN cluster
s1_or_plot <- ggplot(data = s1_output, aes(x = GroupConv.OR, y = -log10(model.pvalue), label = cluster)) +
  theme_bw() + 
  theme(axis.title.x = element_text(size = 9),
        axis.title.y = element_text(size = 9), 
        axis.text.x = element_text(color = "black", size = 9),
        axis.text.y = element_text(color = "black", size = 9),
        plot.title = element_text(hjust = 0.5, size = 12),
        plot.subtitle = element_text(size = 9),
        axis.line.y.left = element_line(color = "black"),
        axis.line.x.bottom = element_line(color = "black"),
        panel.grid = element_blank()) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "#196293") +
  geom_errorbarh(aes(xmin=GroupConv.OR.95pct.ci.lower, xmax=GroupConv.OR.95pct.ci.upper), col = "darkgrey") +
  geom_point(size = 2) + 
  geom_point(data = s1_highlight, 
             aes(x = GroupConv.OR, y = -log10(model.pvalue)), 
             color = "#9D1213",
             size = 2) +
  geom_hline(yintercept = -log10(.05), linetype = "dashed", color = "#9D1213") + 
  scale_x_log10(breaks = c(0.5, 1, 2, 4, 8)) +
  geom_label_repel(size = 3) +
  labs(title = "S1",
       subtitle = "Naive                                                                                           Conv",
       x = "Odds Ratio",
       y = expression("-log"[10]*"(p-value)")) +
  force_panelsizes(rows = unit(3, "in"),
                     cols = unit(4, "in"))

s1_or_plot
```

## Spike 2

```{r, fig.width=9, fig.height=6}
# Get data frame 
s2_df <- s2@meta.data %>%
  select(PTID, nCount_RNA, nCount_ADT, percent.mt, Group, wsnn_res.1)
s2_df <- inner_join(s2_df, metadata, by = "PTID")
s2_df$Group <- factor(s2_df$Group, levels = c("Naive", "Conv"))
s2_df$Sex <- factor(s2_df$Sex, levels = c("Female", "Male"))

# Rescale UMIs
s2_df$nCount_RNA <- scale(log2(s2_df$nCount_RNA))
s2_df$nCount_ADT <- scale(log2(s2_df$nCount_ADT))

# Create design matrix
s2_cluster <- as.character(s2_df$wsnn_res.1)
s2_designmat <- model.matrix(~ s2_cluster + 0, data.frame(cluster = s2_cluster))
s2_df <- cbind(s2_designmat, s2_df)

# Set variables
contrast <- "Group"
random <- "PTID"
fixed <- c("nCount_RNA", "nCount_ADT", "percent.mt", "Age", "Sex")

# Create base model formula
formula_rhs <- paste0(c(paste0(fixed, collapse = " + "),
                      paste0("(1|", random, ")", collapse = " + ")),
                    collapse = " + ")

# Initialize list to store model objects for each cluster
s2_cluster_models <- vector(mode = "list",
                         length = length(attributes(s2_designmat)$dimnames[[2]]))
names(s2_cluster_models) <- attributes(s2_designmat)$dimnames[[2]]

if(rerun_masc) {
  # Run mixed-effects models for each cluster
  system.time({
    for (i in seq_along(colnames(s2_designmat))) {
      test_cluster <- colnames(s2_designmat)[i]
      null_formula <- as.formula(paste0(c(paste0(test_cluster, " ~ 1 + "),
                                          formula_rhs), collapse = ""))
      full_formula <- as.formula(paste0(c(paste0(test_cluster, " ~ ", contrast, " + "),
                                          formula_rhs), collapse = ""))
      
      # Run null and full mixed-effects models
      null_model <- lme4::glmer(formula = null_formula, data = s2_df,
                                family = binomial, nAGQ = 1, verbose = 0,
                                control = glmerControl(optimizer = "bobyqa"))
      full_model <- lme4::glmer(formula = full_formula, data = s2_df,
                                family = binomial, nAGQ = 1, verbose = 0,
                                control = glmerControl(optimizer = "bobyqa"))
      
      # Likelihood ratio test to test fixed effects
      model_lrt <- anova(null_model, full_model)
  
      # Calculate confidence intervals for contrast term beta
      contrast_lvl2 <- paste0(contrast, levels(s2_df[[contrast]])[2])
      contrast_ci <- confint.merMod(full_model, method = "Wald",
                                    parm = contrast_lvl2)
      
      # Save model objects to list
      s2_cluster_models[[i]]$null_model <- null_model
      s2_cluster_models[[i]]$full_model <- full_model
      s2_cluster_models[[i]]$model_lrt <- model_lrt
      s2_cluster_models[[i]]$confint <- contrast_ci
      }  
    })
  
  # Organize results into output dataframe
  s2_output <- data.frame(cluster = attributes(s2_designmat)$dimnames[[2]],
                          size = colSums(s2_designmat))
  s2_output$model.pvalue <- sapply(s2_cluster_models, function(x) x$model_lrt[["Pr(>Chisq)"]][2])
  s2_output$cluster <- s2_output$cluster %>%
    str_replace("s2_cluster", "")
  
  # Get odds ratio for Conv vs. Naive
  s2_output[[paste(contrast_lvl2, "OR", sep = ".")]] <- sapply(s2_cluster_models, function(x) exp(fixef(x$full)[[contrast_lvl2]]))
  s2_output[[paste(contrast_lvl2, "OR", "95pct.ci.lower", sep = ".")]] <- sapply(s2_cluster_models, function(x) exp(x$confint[contrast_lvl2, "2.5 %"]))
  s2_output[[paste(contrast_lvl2, "OR", "95pct.ci.upper", sep = ".")]] <- sapply(s2_cluster_models, function(x) exp(x$confint[contrast_lvl2, "97.5 %"]))
  } else {
  s2_output <- readRDS(file = here::here("out/mixed_models/S2_masc_output.rds"))
}

# Select significant points to highlight 
s2_highlight <- s2_output %>%
  dplyr::filter(-log10(model.pvalue) > -log10(0.05))

# Visualize associations between infection status (Group) and WNN cluster
s2_or_plot <- ggplot(data = s2_output, aes(x = GroupConv.OR, y = -log10(model.pvalue), label = cluster)) +
  theme_bw() + 
  theme(axis.title.x = element_text(size = 9),
        axis.title.y = element_text(size = 9), 
        axis.text.x = element_text(color = "black", size = 9),
        axis.text.y = element_text(color = "black", size = 9),
        plot.title = element_text(hjust = 0.5, size = 12),
        plot.subtitle = element_text(size = 9),
        axis.line.y.left = element_line(color = "black"),
        axis.line.x.bottom = element_line(color = "black"),
        panel.grid = element_blank()) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "#196293") +
  geom_errorbarh(aes(xmin=GroupConv.OR.95pct.ci.lower, xmax=GroupConv.OR.95pct.ci.upper), col = "darkgrey") +
  geom_point(size = 2) + 
  geom_point(data = s2_highlight, 
             aes(x = GroupConv.OR, y = -log10(model.pvalue)), 
             color = "#9D1213",
             size = 2) +
  geom_hline(yintercept = -log10(.05), linetype = "dashed", color = "#9D1213") + 
  scale_x_log10(breaks = c(0.25, 1, 4, 16, 64)) +
  geom_label_repel(size = 3) +
  labs(title = "S2",
       subtitle = "Naive                                                                                          Conv",
       x = "Odds Ratio",
       y = expression("-log"[10]*"(p-value)")) +
  force_panelsizes(rows = unit(3, "in"),
                     cols = unit(4, "in"))

s2_or_plot
```

# Save output

```{r}
if(save_output) {
  saveRDS(s1_output, file = here::here("out/mixed_models/S1_masc_output.rds"))
  
  saveRDS(s2_output, file = here::here("out/mixed_models/S2_masc_output.rds"))
  
  cairo_pdf(file=here::here("out/mixed_models/S1_S2_mixed_models.pdf"), width=5, height=4,
        onefile = TRUE, bg = "transparent", family = "Arial") # default unit is inches, default font is Helvetica.
  print(s1_or_plot)
  print(s2_or_plot)
  dev.off()
}
```
